Automatically proving equivalence by type-safe reflection

### 2016-01-19 paper-JFP-DTP/content/2-basicIdeas.tex
Various fix

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
e1 : (x, y, z : Nat) $\rightarrow$ 
           Expr [x, y, z]
e1 x y z = Plus (Plus (Var FZ) (Var (FS FZ))) 
                (Plus (Var FZ) (Var (FS (FS FZ))))
\end{lstlisting}
\end{center}
\caption{Reflected LHS with the first version of reflected numbers showed in \ref{reflectedNaturalNumbers0}}
\figrule
\end{figure}

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
expr_l : Expr $\Gamma$ x 
         $\rightarrow$ (x' ** (LExpr $\Gamma$ x', x = x'))
expr_l Zero = (_ ** (LZero, Refl))
expr_l (Var i) = (_ ** (LPlus LZero i, Refl))
expr_l (Plus ex ey) = 
  let (x' ** (ex', px)) = expr_l ex in
  let (y' ** (ey', py)) = expr_l ey in
  let (res ** (normRes, Pres)) = plusLExpr ex' ey' in
    (res ** (normRes, rewrite px in (rewrite py in Pres)))
      where 
      plusLExpr : {$\Gamma$ : Vect n Nat} $\rightarrow$ {x, y : Nat} 
            $\rightarrow$ LExpr $\Gamma$ x -> LExpr $\Gamma$ y  
            $\rightarrow$ (z ** (LExpr $\Gamma$ z, x+y=z))
      plusLExpr {x=x} ex LZero =
        (_ ** (ex, rewrite (plusZeroRightNeutral x) in Refl))            
      plusLExpr ex (LPlus e i) =
        let (xRec ** (rec, prfRec)) = plusLExpr ex e in
            (_ ** (LPlus rec i, ?MplusLExpr))

\end{lstlisting}
\end{center}
\caption{Production of the left associative form}
\label{expr_l}
\figrule
\end{figure}
In the case of an addition $Plus\ ex\ ey$, the function $expr\_l$ does the job of normalisation recursively on $ex$ and on $ey$, and then it uses the sub-function $plusLExpr$ to normalise the addition of these two --already normalised-- terms. This sub-function $plusLExpr$ has two kind of simplifications to do. When the second argument is an $LZero$, it simply returns its first arguments along with the justification for this rewriting, which obviously uses $plusZeroRightNeutral$. However, when the second argument is an $LPlus\ e\ i$, it continues recursively by computing $plusLExpr\ ex\ e$, and it finally adds $i$ to it. That had for effect to move the parenthesis on the left, and this treatment is going to be justified during the proof of the meta variable $MplusLExpr$ by the use of $plusAssociative$.

>>>>

\begin{lstlisting}
e1 : (x, y, z : Nat) $\rightarrow$ 
           Expr [x, y, z]
e1 x y z = Plus (Plus (Var FZ) (Var (FS FZ))) 
                (Plus (Var FZ) (Var (FS (FS FZ))))
\end{lstlisting}

@@

\begin{lstlisting}
expr_l : Expr $\Gamma$ x 
         $\rightarrow$ (x' ** (LExpr $\Gamma$ x', x = x'))
expr_l Zero = (_ ** (LZero, Refl))
expr_l (Var i) = (_ ** (LPlus LZero i, Refl))
expr_l (Plus ex ey) = 
  let (x' ** (ex', px)) = expr_l ex in
  let (y' ** (ey', py)) = expr_l ey in
  let (res ** (normRes, Pres)) = plusLExpr ex' ey' in
    (res ** (normRes, rewrite px in (rewrite py in Pres)))
      where 
      plusLExpr : {$\Gamma$ : Vect n Nat} $\rightarrow$ {x, y : Nat} 
            $\rightarrow$ LExpr $\Gamma$ x -> LExpr $\Gamma$ y  
            $\rightarrow$ (z ** (LExpr $\Gamma$ z, x+y=z))
      plusLExpr {x=x} ex LZero =
        (_ ** (ex, rewrite (plusZeroRightNeutral x) in Refl))            
      plusLExpr ex (LPlus e i) =
        let (xRec ** (rec, prfRec)) = plusLExpr ex e in
            (_ ** (LPlus rec i, ?MplusLExpr))

\end{lstlisting}

In the case of an addition $Plus\ ex\ ey$, the function $expr\_l$ does the job of normalisation recursively on $ex$ and on $ey$, and then it uses the sub-function $plusLExpr$ to normalise the addition of these two --already normalised-- terms. This sub-function $plusLExpr$ has two kind of simplifications to do. When the second argument is an $LZero$, it simply returns its first arguments along with the justification for this rewriting, which obviously uses $plusZeroRightNeutral$. However, when the second argument is an $LPlus\ e\ i$, it continues recursively by computing $plusLExpr\ ex\ e$, and it finally adds $i$ to it. That had for effect to move the parenthesis on the left, and this treatment is going to be justified during the proof of the meta variable $MplusLExpr$ by the use of $plusAssociative$.

### 2016-01-19 paper-JFP-DTP/content/3-generalProblem.tex
Various fix

As we did for the very specific problem in section 2, we want to program an automatic reflection mechanism, in order to let the machine build automatically the encodings for us. This time, we want to write a function able to compute the ring encoding for any type that behaves as a ring. Said differently, we've got a carrier type $(c:Type)$, an instance $(p:Ring\ c)$ that says that $c$ behaves as a ring, a context $\Gamma$ of already abstracted variables, and an element $(x:c)$ to encode, and we want to return an $ExprR\ p\ \Gamma\ x$.
We are doing this reflection for any type $c$, so we don't know which constants are inhabiting this type. For this reason, we will need to expect an extra argument --$funReflectCsf$-- of type $(ConstantsRingReflector\ c\ p)$ which is a function able to do the reflection for constants only. We will be able to provide this function for various datatypes, like $Nat$, $Z$ and $List$. But if the user wants to use the reflection mechanism for a new datatype of this own, then he will only have to provide this function of encoding for the constants of his type. He won't have to rebuild the entire reflection function.
A $ConstantsRingReflector$ for a given type $c$ and a given instance $p$ is a function that takes any context $\Gamma$, any element of $c$, and which tries to produce an $(ExprR\ p\ \Gamma\ x)$.

>>>>

As we did for the very specific problem in section 2, we want to program an automatic reflection mechanism, in order to let the machine build automatically the encodings for us. This time, we want to write a function able to compute encodings for any type that behaves as one of our algebraic structure. We will show here the details for Ring, which is one of the most sophisticated structures. So, we've got a carrier type $(c:Type)$, an instance $(p:Ring\ c)$ which says that $c$ behaves as a ring, a context $\Gamma$ of already abstracted variables, and an element $(x:c)$ to encode, and we want to return an $ExprR\ p\ \Gamma\ x$.
We are doing this reflection for any type $c$, so we don't know which constants are inhabiting this type. For this reason, we will need to expect an extra argument --$funReflectCsf$-- of type $(ConstantsRingReflector\ c\ p)$ which is a function able to do the reflection for constants only. We will be able to provide this function for various datatypes, like $Nat$, $Z$ and $List$. But if the user wants to use the reflection mechanism for a new datatype of this own, then he will only have to provide this function of encoding for the constants of his type. He won't have to rebuild the entire reflection mechanism.
A $ConstantsRingReflector$ for a given type $c$ and a given instance $p$ is a function that takes any context $\Gamma$, any element of $c$, and which tries to produce an $(ExprR\ p\ \Gamma\ x)$.

### 2016-01-19 paper-JFP-DTP/content/5-results.tex
Various fix

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
plus (plus c (plus bit (plus v (plus v 0)))) (plus bit (plus v (plus v 0))) 
   = plus v2 (plus (plus (plus v1 v) v) (plus (plus (plus v1 v) v) 0))
\end{lstlisting}
\end{center}
\label{goal}
\figrule
\end{figure}

In a context we're we've got the following induction hypothesis \[ihn:(plus\ (plus\ c\ bit)\ bit1\ =\ plus\ vLsb\ (plus\ vCarry0\ (plus\ vCarry0\ 0)))\]
The goal can't be solved directly by calling the commutative monoid prover, as it uses different variables in its left and right hand side. However, the induction hypothesis $ihn$ --obtained with the recursive call-- creates a link between these variables. Therefore, the solution is simply to use our reduction procedure for commutative monoid four times : two for reducing the LHS and RHS of $ihn$ --that gives a new proof of equality $ihn'$--, and two for reducing the LHS and RHS of the goal. After these four normalisations, it becomes possible to simply rewrite the new induction impothesis $ihn'$ on the current goal.

>>>>

\begin{lstlisting}
plus (plus c (plus bit (plus v (plus v 0)))) (plus bit (plus v (plus v 0))) 
   = plus v2 (plus (plus (plus v1 v) v) (plus (plus (plus v1 v) v) 0))
\end{lstlisting}

In a context we're we've got the following induction hypothesis

\begin{lstlisting}
ihn:(plus (plus c bit) bit1 = plus vLsb (plus vCarry0 (plus vCarry0 0)))
\end{lstlisting}

This goal can't be solved directly by calling the commutative monoid prover, as it uses different variables in its left and right hand side. However, the induction hypothesis $ihn$ --obtained with the recursive call-- creates a link between these variables. Therefore, the solution is simply to use our reduction procedure for commutative monoid four times : two for reducing the LHS and RHS of $ihn$ --that gives a new proof of equality $ihn'$--, and two for reducing the LHS and RHS of the goal. After these four normalisations, it becomes possible to simply rewrite the new induction impothesis $ihn'$ on the current goal.

### 2016-01-19 paper-JFP-DTP/content/6-discussion.tex
Various fix

At the connexion bewteen proof automation and proof by reflection, we can cite SSreflect~\cite{GonthierTuto}, which is an extension of Coq which aims to provide additional formalisation and tactics suitable for long mathematical proofs.
\\

@@

Finally, when we leave the ground of nice mathematical structures (like groups, rings and fields), one can decide to work with various rewriting rules, but in the general case there isn't a complete decision procedure for such problems, because there's usually no normal form. This is where deduction modulo~\cite{Dowek03} ~\cite{DelahayeModulo} starts.

@@

This paper has shown that type-safe reflection techniques can enable to manipulate proofs terms easily. We've been able to build an entire hierarchy of tactics for algebraic structures precisely because the construction of the proof terms has been greatly facilitated by the use this technique. We end up with a hierarchy of reflexive tactics for proving equivalence in semigroups, monoids, commutative monoids, groups, commutative groups, pre semi-ring, semi-ring, rings and commutative rings, without having had to repeat our efforts.

It is even possible to extend a prover in order to build a dedicated one for a specific structure that isn't exactly a well known algebraic structure. We are at the moment investigating the extention of the pre semi-ring prover in order to build a prover for regular expressions. Regular expressions have a union of language playing the role of $+$ (which is associative, commutative, and which has a neutral element $\emptyset$) and a product of language playing the role of $*$, which is associative, distributive over the union, and which admits the empty word $\epsilon$ for neutral element. However, it has also a Kleene $star$ construction and a few properties about it. We currently aim to extend the pre semi-ring prover in order to build a prover for regular expression. Such a prover would be useful for doing formally verified transformations of regular grammars and final states machines for free, which could go up to a formally verified (for free!) parser generator for regular languages.

>>>>

At the connexion bewteen proof automation and proof by reflection, we can cite SSreflect~\cite{GonthierTuto}, which is an extension of Coq which aims to provide additional formalisations and tactics suitable for long mathematical proofs.
\\

@@

Finally, when we leave the ground of nice mathematical structures (like groups, rings and fields), one can decide to work with arbitrary rewriting rules, but in the general case there isn't a complete decision procedure for such problems, because there's usually no normal form. This is where deduction modulo~\cite{Dowek03} ~\cite{DelahayeModulo} starts.

@@

This paper has shown that type-safe reflection techniques can enable to manipulate proofs terms easily. We've been able to build an entire hierarchy of tactics for algebraic structures precisely because the construction of the proof terms has been greatly facilitated by the use of this technique. We end up with a hierarchy of reflexive tactics for proving equivalence in semigroups, monoids, commutative monoids, groups, commutative groups, pre semi-ring, semi-ring, rings and commutative rings, without having had to repeat our efforts.

It is even possible to extend a prover in order to build a dedicated one for a specific structure that isn't exactly a well known algebraic structure. We are at the moment investigating the extention of the pre semi-ring prover in order to build a prover for regular expressions. Regular expressions have a union of language playing the role of $+$ (which is associative, commutative, and which has a neutral element $\emptyset$) and a product of language playing the role of $*$, which is associative, distributive over the union, and which admits the empty word $\epsilon$ for neutral element. However, it has also a Kleene $star$ construction and a few properties about it. We currently aim to extend the pre semi-ring prover in order to build a prover for regular expression. Such a prover would be useful for doing formally-verified-for-free transformations of regular grammars and final states machines, which could go up to a formally verified (for free!) parser generator for regular languages.
